---
title: "Hw05 - Regional Count Data Homework (Local Rates)"
format: html
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

The `nc.rda` file contains information related to SIDS cases in North Carolina. The object includes `nc.sids`, a spatial data frame compatible with the **sf** package with 20 variables measured for 100 counties. It contains data given in Cressie (1991, pp. 386-9), Cressie and Read (1985), and Cressie and Chan (1989) on sudden infant deaths in North Carolina for 1974-78 and 1979-84. The `nc.rda` objects also contains the neighbor list given by Cressie and Chan (1989) omitting self-neighbors (nb89), and the neighbor list given by Cressie and Read (1985) for contiguities (nb85). The `nc.sids` object contains the following variables:

-   `SP_ID`: SpatialPolygons ID
-   `CNTY_ID`: county ID
-   `east`: eastings, county seat, miles, local projection
-   `north`: northings, county seat, miles, local projection
-   `L_id`: Cressie and Read (1985) L index
-   `M_id`: Cressie and Read (1985) M index
-   `names`: County names
-   `AREA`: County polygon areas in degree units
-   `PERIMETER`: County polygon perimeters in degree units
-   `CNTY_` Internal county ID
-   `NAME`: County names
-   `FIPS`: County ID
-   `FIPSNO`: County ID
-   `CRESS_ID`: Cressie papers ID
-   `BIR74`: births, 1974-78
-   `SID74`: SID deaths, 1974-78
-   `NWBIR74`: non-white births, 1974-78
-   `BIR79`: births, 1979-84
-   `SID79`: SID deaths, 1979-84
-   `NWBIR79`: non-white births, 1979-84

# Problem 1

Construct plots of the two neighbor relationships (`nb85` and `nb89`) centered at the approximate centroids of each county (i.e., use the `st_centroids` function to get the approximate centroid of each county for plotting purposes). What differences do you notice in the two plots?

**Solution**

```{r}

library(spdep)
library(sf)
library(smerc)


load(file="nc.rda")

# plot region boundaries from nc.sids
plot(sf::st_geometry(nc.sids$geometry))
# plot neighbors nb85
plot(nb85, coords = st_centroid(st_geometry(nc.sids$geometry)),
     add=TRUE, col="blue", pch = 19, cex = 0.6)


# plot region boundaries from nc.sids
plot(sf::st_geometry(nc.sids$geometry))

# plot neighbors nb89
plot(nb89, coords = st_centroid(st_geometry(nc.sids$geometry)),
     add=TRUE, col="blue", pch = 19, cex = 0.6)

```

> The first plot for the nb85 neighbors appears to have more connections all across the study area with fewer patterns of clustering while the nb89 area has more white space (less connections ar a higher spatial scale) and appears to have some evidence of clustering in the northeast region of our study area.

# Problem 2

## (a)

Create choropleth maps (a plot that colors each region with a value indicating the response level) of the `SID74` and `SID79` variables.

**Solution**

```{r}

plot(nc.sids["SID74"])


plot(nc.sids["SID79"])


```

## (b)

Where do the SIDS deaths tend to be highest? Is there a lot of discrepancy between the plots of the two data sets, or is the overall pattern similar?

**Solution**

> In general, it appears the intensity of SIDs deaths located in the center bottom region, with a slight translation of SIDs deaths from right to left between the SID74 to the SID79 plot.

## (c)

Why must we be cautious in concluding the brighter regions are most likely to be hotspots/clusters of cases? What might be a better way of displaying the data?

**Solution**

> The brighter regions may have more total number of cases but also have a much higher population or number of births. A better way to display the data would be to show the intensity in that county relative to the county's total population or total birth count.

# Problem 3

Use the CEPP method to identify potential clusters for the `SID74` variable. Note that the "population" associated with each region should be the associated number of births during the respective time period, provided in `BIR74`. Use a significance level of 0.01 and $n^*=$ 2000, 20000, and 100000. Use `east` and `north` (the x and y coordinates) to define the centroid associated with each region. Do not use the centroids you obtained from the `st_centroids` function.

```{r}

# set parameters
coords <- cbind(nc.sids$east,nc.sids$north)
cases = nc.sids$SID79
pop = nc.sids$BIR74


# Identify potential clusters at n* = 2000
cepp2k = cepp.test(coords = coords,
                     cases = cases,
                     pop = pop,
                     nstar = 2000,
                     alpha = 0.01)

# summary of significant clusters
summary(cepp2k)



# Identify potential clusters at n* = 20000
cepp20k = cepp.test(coords = coords,
                     cases = cases,
                     pop = pop,
                     nstar = 20000,
                     alpha = 0.01)

# summary of significant clusters
summary(cepp20k)




# Identify potential clusters at n* = 100,000
cepp100k = cepp.test(coords = coords,
                     cases = cases,
                     pop = pop,
                     nstar = 100000,
                     alpha = 0.01)

# summary of significant clusters
summary(cepp100k)

```

## (a)

State the null and alternative hypothesis test in the context of the problem for a generic $n^*$.

**Solution**

> Null: There is no window with n\* persons at risk that has significantly more cases than what would be expected under CRH.
>
> Alternative: There is at least one window with n\* persons at risk that has significantly more cases than would be expected under CRH.

## (b)

Interpret the overall results in the context of the problem.

**Solution**

> We see evidence of clustering of intensity of cases relative to what would be expected under CRH for n\* = 2,000 and n\* = 20,000 at the alpha = 0.01 level. Specifically, we can conclude that the intensity in the identified regions is not consistent with the constant risk hypothesis at the n\* = 2,000 persons at risk and n\* = 20,000 persons at risk.

## (c)

Plot the results on the map of North Carolina counties. Don't worry too much about clearly identifying the clusters (cluster A, B, C, etc.). Look at the overarching patterns of the plots. Are the clusters in roughly the same areas? Are the clusters roughly the same size (in terms of number of regions?) Explain.

**Solution**

```{r}

# plot of most likely clusters for n* = 2000
plot(sf::st_geometry(nc.sids), border = "grey60", axes = TRUE,
     col = color.clusters(cepp2k))
legend("topright", legend = c("n* = 2000"))


# plot of most likely clusters for n* = 20000
plot(sf::st_geometry(nc.sids), border = "grey60", axes = TRUE,
     col = color.clusters(cepp20k))
legend("topright", legend = c("n* = 20,000"))


# plot of most likely clusters for n* = 100,000
plot(sf::st_geometry(nc.sids), border = "grey60", axes = TRUE,
     col = color.clusters(cepp100k))
legend("topright", legend = c("n* = 100,000"))

```

> The clusters appear to be in the same relative region (lower central region) however the number of regions for each plot more than doubles from the previous - this makes sense because as we increase the threshold for n\* (number of persons at risk identified in the window) the more regions will be in the window with that many persons at risk.

# Problem 4

Use the Besag-Newell method to identify potential clusters for the `SID74` variable. Note that the "population" associated with each region should be the associated number of births during the respective time period, provided in `BIR74`. Use a significance level of 0.01 and $c^*=$ 6, 12, and 25. Use `east` and `north` (the x and y coordinates) to define the centroid associated with each region. Do not use the centroids you obtained from the `st_centroids` function.

```{r}

# set parameters
coords <- cbind(nc.sids$east,nc.sids$north)
cases = nc.sids$SID79
pop = nc.sids$BIR74

bn6 = bn.test(coords = coords,
              cases = cases,
              pop = pop,
              cstar = 6,
              alpha = 0.01)

summary(bn6)
clusters(bn6)

bn12 = bn.test(coords = coords,
              cases = cases,
              pop = pop,
              cstar = 12,
              alpha = 0.01)

summary(bn12)
clusters(bn12)


bn25 = bn.test(coords = coords,
              cases = cases,
              pop = pop,
              cstar = 25,
              alpha = 0.01)

summary(bn25)
clusters(bn25)
```

## (a)

State the null and alternative hypothesis test in the context of the problem for a generic $c^*$.

**Solution**

> Null: The most compact window (population wise) with at least c\* cases is not significantly more compact than what is expected under the CRH.
>
> Alternative: The most compact window (in terms of population size) with at least c\* cases is significantly more compact than what is expected under the CRH.

## (b)

Interpret the overall results in the context of the problem.

**Solution**

> Based on the besag-newell test, we do not find evidence that the most compact window with c\* cases is more compact than it would be under CRH for any value of c\*. In other words, we do not find evidence of clustering of SIDs deaths more than would be expected under CRH.

## (c)

Plot the results on the map of North Carolina counties. Don't worry too much about clearly identifying the clusters (cluster A, B, C, etc.). Look at the overarching patterns of the plots. Are the clusters in roughly the same areas? Are the clusters roughly the same size (in terms of number of regions?) Explain.

**Solution**

```{r}

library(RColorBrewer) 

display.brewer.all(type = "qual", colorblindFriendly = TRUE)
mycol = brewer.pal(3, "Dark2")

# create vector of colors to show results

# default is white (no clustering)
nccol = rep("white", nrow(nc.sids))
# the most likely cluster locations for cstar = 6 are green
nccol[bn6$clusters[[1]]$locids] = mycol[1]
# the most likely cluster locations for cstar = 12 are orange
nccol[bn12$clusters[[1]]$locids] = mycol[2]
# the most likely cluster locations for cstar = 25 are violet
nccol[bn25$clusters[[1]]$locids] = mycol[3]


# plot most likely clusters
plot(st_geometry(nc.sids), border="grey60", axes = TRUE, col = nccol)
legend("topleft",
       legend = c("Cluster k = 6 (Green)", "Cluster k = 12 (Orange)",
                  "Cluster k = 25 (Violet)"),
       lwd = 10, col = nccol)
```

> The clusters for k=6 and k=12 are the same size, the cluster for k=25 has an additional 2 regions. The first two clusters are in the same lower central area of the study area while the k=25 cluster is in the top right corner of the study area.

# Problem 5

Use the Poisson spatial scan statistic method under the CRH to identify potential clusters for the `SID74` variable. Note that the "population" associated with each region should be the associated number of births during the respective time period, provided in `BIR74`, respectively. Use a significance level of 0.10. Use two different population upper bounds: 0.1 and 0.5. Use `east` and `north` (the x and y coordinates) to define the centroid associated with each region. Do not use the centroids you obtained from the `st_centroids` function.

```{r}

# set parameters
coords <- cbind(nc.sids$east,nc.sids$north)
cases = nc.sids$SID79
pop = nc.sids$BIR74


# expected number of cases
e = sum(cases)/sum(pop) * pop
# apply circular scan method
scan = scan.test(coords = coords,
                 cases = cases,
                 pop = pop,
                 ex = e,
                 nsim = 999,
                 alpha  = 0.1)

# results from the test are available in
summary(scan)
# cluster information
clusters(scan)
```

## (a)

State the null and alternative hypotheses in the context of the problem.

**Solution**

> Null: The most likely cluster of cases (local rate in cluster relative to outside of cluster) is consistent with what is expected under the CRH.
>
> Alternative: The most likely cluster (in terms of local rate of cases vs the rate outside of the cluster) is more extreme than what is expected under CRH.

## (b)

Interpret the overall results in the context of the problem.

**Solution**

> At the alpha = 0.1 level, we can reject the null hypothesis and conclude that there is evidence that the most likely cluster identified is more extreme in intensity of cases within the cluster relative to intensity of cases outside of the cluster than what would be expected under CRH.

## (c)

Plot the results on the map of North Carolina counties. Don't worry too much about clearly identifying the clusters (cluster A, B, C, etc.). Look at the overarching patterns of the plots. Are the clusters in roughly the same areas? Are the clusters roughly the same size (in terms of number of regions?) Explain.

**Solution**

```{r}


# need to color 1 clusters
mycol = grDevices::hcl.colors(1)

plot(sf::st_geometry(nc.sids), border="grey60", axes=TRUE,
     col = color.clusters(scan, col = mycol))
legend("topright", legend = c("Cluster A"),
       lwd = 10, col = mycol)

```

> The scan statistic only returned a single cluster. The location of this cluster is similarly in the lower central region, as most clusters we plotted with previous methods were.

# Problem 6

How do the plots for the CEPP, Besag-Newell, and scan methods compare to one another? Do you believe there is a cluster (or clusters) of SIDS cases present in the data? Explain.

**Solution**

> Most of the statistically significant evidence of clustering identified the clusters in the same general region - somewhere in the lower central area of our study area. This leads me to believe that there is significant evidence of clusters of SIDs cases, or at least of one cluster, and the location of that cluster is likely in the lower central region of our study area.
